Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting

### 2025-04-24 code/analysis-code/Statistical_Analysis.qmd
Project4

```{r}
library(ggplot2)

ggplot(promo_data_clean, aes(x = total_malaria_episodes, fill = study_arm)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "dodge") +
  labs(title = "Distribution of Total Malaria Episodes by IPTp Treatment Arm",
       x = "Total Malaria Episodes",
       y = "Frequency",
       fill = "Treatment Arm") +
  theme_minimal()

@@

@@

ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1) +
  labs(title = "ROC Curves: ML Models for Adverse Outcome Prediction",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```
Figure X displays the ROC curves for logistic regression (red), random forest (green), and gradient boosting (blue) on the held‑out test set. All three curves lie close to the diagonal, reflecting poor discrimination: none of the models reliably separates pregnancies with versus without adverse outcomes. Of the three, logistic regression achieved the highest AUC (0.48), followed by XGBoost (0.44) and random forest (0.42), but these differences are marginal and all fall below the 0.50 threshold for random performance. Collectively, these results underscore that, with the current predictor set, more informative features or alternative modeling strategies will be necessary to improve predictive accuracy.

@@

@@

```{r ml_improved_setup, message=FALSE, warning=FALSE}
# 0) Load all needed packages
library(tidymodels)  # recipes, parsnip, workflows, tune, yardstick, dials
library(themis)      # for step_smote()
```

@@

```{r}
library(themis)

imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  # standard preprocessing
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # adaptive synthetic sampling
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

>>>>

```{r}
ggplot(promo_data_clean, aes(x = total_malaria_episodes, fill = study_arm)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "dodge") +
  labs(title = "Distribution of Total Malaria Episodes by IPTp Treatment Arm",
       x = "Total Malaria Episodes",
       y = "Frequency",
       fill = "Treatment Arm") +
  theme_minimal()

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Subset to women under 25 and define outcome. Logistic Regression
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25) %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(
      preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0
    )
  )

# Fit logistic regression model
model_gravidity <- glm(
  adverse_birth_outcome ~ gravidity + total_malaria_episodes + 
    study_arm + education_level,
  family = binomial(link = "logit"),
  data = promo_data_young
)

# Load broom for tidy output
library(broom)

# Display model summary as odds ratios with 95% CI
tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

@@

ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1) +
  labs(title = "ROC Curves: ML Models for Adverse Outcome Prediction",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```
Figure displays the ROC curves for logistic regression (red), random forest (green), and gradient boosting (blue) on the held‑out test set. All three curves lie close to the diagonal, reflecting poor discrimination: none of the models reliably separates pregnancies with versus without adverse outcomes. Of the three, logistic regression achieved the highest AUC (0.48), followed by XGBoost (0.44) and random forest (0.42), but these differences are marginal and all fall below the 0.50 threshold for random performance. Collectively, these results underscore that, with the current predictor set, more informative features or alternative modeling strategies will be necessary to improve predictive accuracy.

@@

**Include**
```{r}
ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1) +
  labs(title = "ROC Curves: ML Models for Adverse Outcome Prediction",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

# Model Comparison Table (AUC values)

```{r}
tibble::tibble(
  Model = c("Logistic Regression", "Random Forest", "XGBoost", "Elastic Net"),
  AUC = c(0.48, 0.42, 0.44, 0.49)
)

@@

@@

```{r}
imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  # standard preprocessing
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # adaptive synthetic sampling
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

### 2025-04-24 products/manuscript/Manuscript.qmd
Project4

**Table 1: Baseline Characteristics of Study Participants by IPTp Treatment Arm**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create table data
table_data <- tribble(
  ~Characteristic,                           ~Subcategory,               ~`DP (N=4121)`,       ~`SP (N=3701)`,
  "Age (years)",                             "",                         "25.0 (5.0)",         "25.0 (4.9)",
  "Gestational Age (weeks)",                 "",                         "15.24 (1.97)",       "14.97 (1.94)",
  "Maternal Education Level (%)",            "",                         "",                   "",
  "",                                        "Primary",                  "262 (66.7%)",        "258 (66.4%)",
  "",                                        "Tertiary",                 "65 (16.5%)",         "61 (15.7%)",
  "",                                        "University",               "66 (16.8%)",         "70 (18.0%)",
  "Gravidity",                               "",                         "",                   "",
  "",                                        "1",                        "91 (22%)",           "56 (15%)",
  "",                                        "2–3",                      "167 (41%)",          "166 (45%)",
  "",                                        "≥4",                       "154 (37%)",          "148 (40%)",
  "Parity",                                  "",                         "",                   "",
  "",                                        "0",                        "79 (19%)",           "77 (21%)",
  "",                                        "1–2",                      "175 (42%)",          "158 (43%)",
  "",                                        "≥3",                       "158 (38%)",          "135 (36%)",
  "Total Malaria Episodes",                  "",                         "",                   "",
  "",                                        "1",                        "245 (59%)",          "216 (58%)",
  "",                                        "2–3",                      "146 (35%)",          "127 (34%)",
  "",                                        "≥4",                       "21 (5.1%)",          "27 (7.3%)",
  "Total Malaria Episodes During Pregnancy", "",                         "",                   "",
  "",                                        "1",                        "179 (43%)",          "165 (45%)",
  "",                                        "2–3",                      "138 (33%)",          "120 (32%)",
  "",                                        "≥4",                       "95 (23%)",           "85 (23%)",
  "Malaria Infection Rate During Pregnancy", "",                         "1.01 (0.58)",        "1.05 (0.58)",
  "Placental Malaria (Rogerson Criteria)",   "",                         "75 (18%)",           "72 (19%)",
  "Preterm Births Count",                    "",                         "",                   "",
  "",                                        "1",                        "391 (95%)",          "348 (94%)",
  "",                                        "2",                        "21 (5.1%)",          "22 (5.9%)",
  "Stillbirth bin",                          "",                         "37 (9.0%)",          "50 (14%)",
  "Birthweight",                             "",                         "2.79 (0.64)",        "2.81 (0.62)",
  "History of Previous Malaria Episodes (%)", "",                        "40 (10%)",           "35 (9%)"
)

kable(
  table_data,
  format = "pandoc",
  caption = ""
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

**Table 4: Interaction Between Malaria Exposure and IPTp Treatment Arm in Predicting Adverse Birth Outcomes**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#library(broom)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)

# Tidy the model output with exponentiated coefficients (odds ratios)
tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)

# Remove the intercept row and recode variable names using case_when()
tidy_model_filtered <- tidy_model %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "age_at_enrollment_years" ~ "Age at Enrollment (years)",
      term == "gravidity" ~ "Gravidity",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      term == "total_malaria_episodes:study_armSP" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ term
    )
  )

# Round key values for clarity
tidy_model_filtered <- tidy_model_filtered %>%
  mutate(
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )

# Render the table with a blank header for the first column
kable(
  tidy_model_filtered,
  format = "pandoc",
  caption = "",
  col.names = c("", "Odds Ratio", "Std. Error", "z value", "p-value", "95% CI Lower", "95% CI Upper")
) %>% 
  kable_styling(full_width = FALSE, position = "center")

**Table 5:  Regression Diagnostics for IPTp Treatment Effect**
```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Read in the cleaned data (ensure adverse_birth_outcome is defined as needed)
promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

# Create composite adverse outcome if not already defined
promo_data_clean <- promo_data_clean %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

# Fit Model 1 (without interaction)
model1 <- glm(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"), data = promo_data_clean)

# Fit Model 2 (with interaction)
model2 <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"), data = promo_data_clean)

# Generate Analysis of Deviance Table (Likelihood Ratio Test)
lrt <- anova(model1, model2, test = "LRT")
lrt_df <- as.data.frame(lrt)
lrt_df$Model <- c("No Interaction", "With Interaction")
lrt_df <- lrt_df %>% 
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)
colnames(lrt_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

# Render Table X: Likelihood Ratio Test
kable(
  lrt_df,
  format = "pandoc",
  caption = ""
) %>% 
  kable_styling(full_width = FALSE, position = "center")

# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

# Render Table Y: VIF Diagnostics
kable(
  vif_df,
  format = "pandoc",
  caption = ""
) %>% 
  kable_styling(full_width = FALSE, position = "center")

**Table 6: Evaluating Effect Modification by IPTp Treatment on Malaria-Related Adverse Birth Outcomes Through Model Comparison**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("C:/Users/ajose35/Desktop/MADA-course/Joseph-MADA-project/Images/Table- Comparing Models .png")

**Table 7: "Incidence Rate Ratios for Malaria Episode Occurrence among Pregnant Women**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Table 4: Incidence Rate Ratios for Malaria Episode Occurrence among Pregnant Women
table_model_refined <- data.frame(
  Variable = c("Intercept", 
               "Treatment Arm (SP vs. DP)", 
               "Age at Enrollment (years)", 
               "Gravidity", 
               "Education Level: Secondary", 
               "Education Level: Tertiary", 
               "Education Level: University"),
  IRR = c(1.6081682, 1.0552044, 0.9942737, 1.0095680, 0.9884423, 0.8900707, 1.0363703),
  `95% CI Lower` = c(1.1297705, 0.9382987, 0.9825662, 0.9675647, 0.8348897, 0.7505115, 0.8805393),
  `95% CI Upper` = c(2.280714, 1.186523, 1.006133, 1.053391, 1.170522, 1.055600, 1.220799),
  `p-value` = c(0.008021017, 0.369364418, 0.342231948, 0.660449862, 0.892661838, 0.180547176, 0.668035030)
)

# Optionally, display the table using knitr::kable for a nicer format
knitr::kable(table_model_refined, caption = "")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Fit the logistic regression model for women under 25
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25) %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

model_gravidity <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + 
                         study_arm + education_level,
                       family = binomial(link = "logit"),
                       data = promo_data_young)

**Table 8: Adjusted Odds Ratios for Adverse Birth Outcomes Among Young Pregnant Women (<25 Years)**
```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Tidy the model output, exponentiating coefficients to get odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

# Create a summary table without the intercept and with neat variable names
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "gravidity" ~ "Gravidity",
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      TRUE ~ term
    ),
    `Odds Ratio (95% CI)` = paste0(round(estimate, 2),
                                   " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  select(term, `Odds Ratio (95% CI)`, p.value) %>%
  rename(Variable = term, `p-value` = p.value)

# Print the table using kable
result_table %>%
  kable(format = "markdown", caption = "")

@@

>>>>

**Table 1: Baseline Characteristics of Study Participants by IPTp Treatment Arm**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics(here::here("Images", "Table 1. Baseline Characteristics .png"))

@@

**Table 4: Interaction Between Malaria Exposure and IPTp Treatment Arm in Predicting Adverse Birth Outcomes**

```{r,echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics(here::here("Images", "Table4.Interaction Malaria vsIPTP.png"))

@@

**Table 5: Adjusted Odds Ratios for Adverse Birth Outcomes Among Young Pregnant Women (<25 Years)**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics(here::here("Images", "Adjusted OR for Adverse Birth Outcomes Women  less 25 Years.png"))

@@

**Figure 4 : Comparative Discriminative Performance of ML Models for Predicting Adverse Birth Outcomes**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
roc_lr  <- roc_curve(logistic_preds, truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "Logistic")
roc_rf  <- roc_curve(rf_preds,       truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "RF")
roc_xgb <- roc_curve(xgb_preds,     truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "XGBoost")

roc_all <- bind_rows(roc_lr, roc_rf, roc_xgb)

ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1) +
  labs(title = "",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```

**Table 6:  Discriminative Performance (AUC) of Machine Learning Models Predicting Adverse Birth Outcomes Under Different IPTp Regimens**
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics(here::here("Images", "Table6.Discriminative Performance (AUC) of ML Models.png"))
